Smooth and Persistent Forecasts of German GDP:\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}

### 2025-08-15 Archive IWH Simon paper/12.08.2025/paper.tex
up

\begin{abstract} %max.
This paper explores the trade-off between informational efficiency and forecast smoothness in economic forecasting. Forecasts that minimize mean squared forecast errors often exhibit excessive volatility, limiting their practical applicability. We propose the Multivariate Smooth Sign Accuracy (M-SSA) framework as a solution, extracting smoothed components from leading indicators. This enhances the signal-to-noise ratio and controls forecast volatility and timing. Applied to German GDP growth, our method yields smoothed forecasts that improve the forecasting accuracy of traditional models, particularly over medium-term horizons. While smoother forecasts tend to lag slightly, this drawback can be mitigated by adjusting the forecast horizon. These findings highlight the practicality of the M-SSA framework for both forecasters and policymakers, especially in settings where updating forecasts or policy adjustments are costly.

@@

\bigskip \vspace{10pt}
\textbf{Keywords:} Forecast Smoothing, Smooth Sign Accuracy, Time Series Filtering 
\newline
\textbf{JEL Classification:} C53, E37, E66
%C53 – Forecasting and Prediction Methods; Simulation Methods
%E37 – Forecasting and Simulation: Models and Applications
%E66 – General Outlook and Conditions: Forecasts and Simulations
\thispagestyle{empty}

@@

Economic forecasters face a dilemma: while informational efficiency requires that forecasts incorporate all available data, doing so often results in forecasts that are excessively volatile.
 %Volatile forecasts may be a sign that they efficiently incorporate all available information. 
 In addition, forecasts may reflect excessive ``noise'' due to overfitting. Further, forecast users (including public and private decision makers) may face constraints in their speed of adjustment to changing forecasts. For example, fixed annual budgets may constrain spending, or transaction costs may limit changes in investments. Others may attempt to smooth their reactions for reasons of signalling. Important examples include dividend smoothing by corporate managers and interest rate smoothing by central banks.

\cite{Wildi2024,Wildi2025} and \cite{McElroy2019,McElroy2020} propose novel methods that optimally smooth forecasts by controlling the expected frequency of sign changes. The resulting Smooth Sign Accuracy (SSA) framework encompasses efficiently the mean-square error (MSE) as a special case and produces increasingly smooth forecasts as changes in the forecast direction (sign) are increasingly penalized.

@@

\subsection{Dependence}
To examine the joint dynamics of our variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \ref{sec:mSSA} and \ref{sec:mSSA_component}.

@@

We then estimated a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}.  This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package.  This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The figures \ref{ma_inv_BIP} and \ref{ma_inv_BIP_var3_en} show the implied responses of GDP to shocks to each indicator variable.\footnote{Shown are the MA coefficients for GDP from the vector moving-average (VMA) models obtained by inversion of the VAR(1) and VAR(3) representations respectively (i.e. the impulse-response functions (IRFs) of GDP to the correlated reduced-form shocks.)}

@@

% where $\lambda_1$ and $\lambda_2$ are non-negative tuning parameters, and $X_i$ represents the predictor variables.
% This formulation encompasses the least-squares estimator as the special case where $\lambda_1 = 0$ and the LASSO as the special case where $\lambda_2 = 0$. Unlike the lasso, which may arbitrarily select correlated predictors, the elastic net encourages grouping of strongly correlated variables. The combined penalties stabilize estimation in cases with high-dimensional predictors.
%
%
%

\begin{figure}[H]
    \begin{center}
        \includegraphics[height=3in, width=6in]{./Figures/ma_inv_BIP.pdf}
        \caption{VMA coefficients for GDP implied by VAR(1)\\Estimation with pandemic (left) and without pandemic (right).
        \label{ma_inv_BIP}}\end{center}
\end{figure}

\begin{figure}[H]
    \begin{center}
        \includegraphics[height=3in, width=6in]{./Figures/ma_inv_BIP_var3_en.pdf}
        \caption{VMA coefficients for GDP implied by VAR(3)\\Estimation with pandemic (left) and without pandemic (right).
        \label{ma_inv_BIP_var3_en}}\end{center}
\end{figure}

@@

In this section, we demonstrate how pre-filtering indicator variables may improve the performance of simple forecasts by removing noise. We then demonstrate how in-sample forecast performance improves when we ``de-noise'' the indicators prior to projection with simple smoothing filters.\footnote{The smoothing filters are one-sided (concurrent) filters of length 31, therefore, estimation of both forecasting models omits the first 30 observations.}

@@

\begin{equation}
    u_{t+h} = \boldsymbol{\beta'} \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $u_{t+h}$ is the quarterly growth rate of GDP for forecast horizon $h = \{0, 1, 2, ... \}$
    \item $\mathbf{X}_t \equiv [1,u_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $u_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\boldsymbol{\beta}}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_all.pdf}
        %\includegraphics[height=3in, width=4.5in]{./Figures/direct_wc_all.pdf}
        \caption{Direct forecasts (red lines) based on all indicators and full sample information (without pandemic.)\\
        Target (black line): GDP left-shifted  by 0 - 3 quarters.
        \label{direct_wc}}
    \end{center}
\end{figure}

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_financial_crisis.pdf}
        \caption{Direct forecast across the financial crisis:  regression based on all indicators and full sample information, without pandemic.
        \label{direct_wc_financial_crisis}}
    \end{center}
\end{figure}

The results indicate that as the forecast horizon $h$ increases, forecast become less volatile and their turning points (i.e., peaks and troughs) become increasingly right-shifted (lagged) relative to the target GDP. Additionally, the p-values of the HAC-adjusted Wald test precipitously increase from $h=2$ to $h=3$ (see Table \eqref{tab:pvaluedhp}), implying that the indicators become statistically insignificant for forecast horizons of more than two quarters.\footnote{ 
These findings are confirmed by the relative root mean-square errors comparing direct forecasts against the mean benchmark, see Table \ref{rRMSE_mSSA_direct_mean_without_covid8} in the Appendix.}

@@

%\begin{table}[ht]
%\centering
%\begin{tabular}{rrrrr}
%  \hline
% & h=3 & h=4 & h=5 & h=6 \\ 
%  \hline
%rRMSE & 0.962 & 0.967 & 0.990 & 0.967 \\ 
%   \hline
%\end{tabular}
%\caption{Relative root mean-square error of HP-C against classic direct forecast for $h\geq %3$.\\Values $<1$ imply  HP-C has smaller RMSE.  } 
%\label{tab:rRMSE}
%\end{table}
\bigskip
\begin{table}[h]
\caption{p-values for $H_0: \boldsymbol{\beta} = 0$ in \eqref{eq:direct_forecast} based on HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}

@@

\begin{figure}[h]
    \begin{center}
%    \textcolor{blue}{\textbf{Notice that the title on top right panel is incorrect!}}
        \includegraphics[height=4in, width=5.5in]{./Figures/hp_160.pdf}
        % \caption{Two-sided HP(160) (top left) and one-sided (concurrent) HP-C filters (top right) with corresponding amplitude functions (bottom).
        \caption{HP$(\lambda=160)$ filters\\
        Left: HP (Two-sided, symmetric) \qquad Right: HP-C (One-sided, concurrent)\\
        Top Row: Filter Weights \qquad \qquad \quad Bottom Row: Amplitude Functions.
        \label{hp_160}}
    \end{center}
\end{figure}

@@

Figures \ref{direct_hp_forecasts} and \ref{direct_hp_forecasts_financial_crisis} provide results comparable to those of Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} but now using indicators smoothed with the one-sided HP(160), (denoted as the HP-C, for \textit{concurrent}) filter.  The fitted values lag GDP slightly less and track turning points somewhat more quickly. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients ${\boldsymbol{\beta}} = 0$. The use of filtered indicators improves their statistical significance at forecast horizons $h > 2$ quarters. 
%Figures \ref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \ref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic (unfiltered) direct forecasts in the latter figure %in Fig.\eqref{direct_wc_financial_crisis} 
% indicates that the new predictors are slightly less retarded and track the turning points more promptly (`faster'). 
% The filtered approach also exhibits superior statistical properties. As shown in Table \ref{tab:f_stats}, 
%The p-values of the F-statistics for the HP-based forecasts are smaller than for the direct forecasts specifically at $h=3$, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators).

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.85\textwidth]{./Figures/direct_hp_forecasts.pdf}
        \caption{Direct HP forecasts: entire data set with exclusion of the pandemic.
        \label{direct_hp_forecasts}}
    \end{center}
\end{figure}

@@

The application of the one-sided HP-C filter to the indicators yields some measurable improvements in forecast performance. However, 
%its univariate nature excludes additional information such as that provided by leading indicators. 
the noise suppression capability of the one-sided filter is compromised, as explained by \cite{Wildi2025} and illustrated by the high-frequency leakage of the amplitude function in Figure \eqref{hp_160}. To address these limitations, we turn to the Multivariate Smooth Sign Accuracy (M-SSA) framework proposed by \cite{Wildi2025}.

@@

Let $\mathbf{x}_t$ be a time series representing the data, with observations $(x_{t},...,x_{1})'$. Let $\mathbf{z}_t=(z_{t},...,z_1)'$  denote a target series, which depends on $\mathbf{x}_t$ as well as on unknown future $x_{t+k}$, $k>0$ (and possibly unknown past $x_k$, $k\leq 0$). For simplicity, we assume stationarity of all time series involved. In our current framwork, $z_t$ corresponds to the output of a two-sided Hodrick–Prescott filter with smoothing parameter $\lambda = 160$ applied to the doubly infinite sequence $x_k$, $k\in\mathbb{Z}$. The finite sample $\mathbf{x}_t$ is the portion effectively observed and represents the log-differenced GDP (the dependence on $x_0, x_{-1},...$ is neglected because the HP filter weights decay rapidly to zero; see Fig. \ref{hp_160}). The prediction task comprises tracking $z_{t}$  using the (causal) predictor $y_{t}=\sum_{k=0}^{L-1} b_{k}x_{t-k}=\mathbf{b}'\mathbf{x}_t$, where the filter $\mathbf{b}$ with weights $b_0,...,b_{L-1}'$ (actually of length $L=t$) is available for optimization purposes.

@@

Traditional predictors often yield an excessive number of false sign changes due to leakage of high-frequency components (high-frequency leakage) in the predicted series, as discussed above (see Figure \ref{hp_160}). This noisiness is particularly problematic near the onset or termination of economic crises (recessions), where the target fluctuates around zero; reliable tracking of sign changes would facilitate real-time assessment of the occurrence and timing of such events. 
%as illustrated by the amplitude function in Figure \ref{hp_160} (bottom right panel) and further elaborated upon by \cite{Wildi2025}. 
 \cite{Wildi2025} shows that the SSA criterion helps control this phenomenon (see Fig.\ref{mssa_msse_zc} for illustration).\\ Using the dual formulation of the criterion \eqref{critssaht}, the author demonstrates that the SSA solution yields the \emph{lowest} zero-crossing rate among all (linear) predictors with equivalent target correlation. 
The concept can be extended to a multivariate framework, denoted M-SSA, by generalizing the correlation functions as detailed in the next section. As we shall see, controlling the zero-crossing rate within the M-SSA framework will have a quantifiable effect on the hit rate and false positive rate, thereby explicitly addressing the detection of present (nowcasting) and future (forecasting) economic contractions and expansions.\\

>>>>

\begin{abstract} %max.
Forecasts that minimize mean squared forecast error (MSE) often exhibit excessive volatility, limiting their practical applicability. We address this accuracy–smoothness trade-off by introducing a Multivariate Smooth Sign Accuracy (M-SSA) framework, which extracts smoothed components from leading indicators to enhance the signal-to-noise ratio and control the forecast volatility and timing. Applied to quarterly German GDP growth, our method yields smoothed forecasts that improve the forecasting accuracy of traditional models, particularly over medium-term horizons. While smoother forecasts tend to lag slightly in turning points, adjusting the forecast horizon offsets this effect. These findings highlight the practicality of the M-SSA framework for both forecasters and policymakers, especially in settings where forecast revisions or policy adjustments are costly.

@@

\bigskip \vspace{10pt}
\textbf{Keywords:} Forecast Smoothing, Smooth Sign Accuracy, Time-Series Filtering 
\newline
\textbf{JEL Classification:} C53, E37, E66
%C53 – Forecasting and Prediction Methods; Simulation Methods
%E37 – Forecasting and Simulation: Models and Applications
%E66 – General Outlook and Conditions: Forecasts and Simulations
\thispagestyle{empty}

@@

Economic forecasters face a dilemma: while informational efficiency requires that forecasts incorporate all available data, doing so often results in forecasts that are excessively volatile.
 %Volatile forecasts may be a sign that they efficiently incorporate all available information. 
 In addition, forecasts may reflect excessive ``noise'' due to overfitting. At the same time, forecast users (including public and private decision makers) may face constraints in their speed of adjustment to changing forecasts. For example, fixed annual budgets may constrain spending, or transaction costs may limit changes in investments. Others may attempt to smooth their reactions for reasons of signalling. Important examples include dividend smoothing by corporate managers and interest rate smoothing by central banks.

Recent work by \cite{Wildi2024,Wildi2025} and \cite{McElroy2019,McElroy2020} formalipropose novel methods that optimally smooth forecasts by controlling the expected frequency of sign changes. The resulting Smooth Sign Accuracy (SSA) framework encompasses efficiently the mean-square error (MSE) as a special case and produces increasingly smooth forecasts as changes in the forecast direction (sign) are increasingly penalized.

@@

\subsection{Dependence}
To examine the joint dynamics of our selected variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \ref{sec:mSSA} and \ref{sec:mSSA_component}.

@@

We then estimated a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}.  This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package. This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The Figures \ref{ma_inv_BIP} and \ref{ma_inv_BIP_var3_en} show the implied responses of GDP to shocks to each indicator variable.\footnote{Shown are the MA coefficients for GDP from the vector moving-average (VMA) models obtained by inversion of the VAR(1) and VAR(3) representations respectively (i.e. the impulse-response functions (IRFs) of GDP to the correlated reduced-form shocks).}

@@

% where $\lambda_1$ and $\lambda_2$ are non-negative tuning parameters, and $X_i$ represents the predictor variables.
% This formulation encompasses the least-squares estimator as the special case where $\lambda_1 = 0$ and the LASSO as the special case where $\lambda_2 = 0$. Unlike the lasso, which may arbitrarily select correlated predictors, the elastic net encourages grouping of strongly correlated variables. The combined penalties stabilize estimation in cases with high-dimensional predictors.
%
%

\begin{figure}[htpb]
    \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
            \includegraphics[height=3in, width=6in]{./Figures/ma_inv_BIP.pdf}
            \caption{implied by VAR(1) \label{ma_inv_BIP}}
        \end{center}
    \end{subfigure}
    \\
    \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
            \includegraphics[height=3in, width=6in]{./Figures/ma_inv_BIP_var3_en.pdf}
            \caption{implied by VAR(3)\label{ma_inv_BIP_var3_en}}
        \end{center}
    \end{subfigure}
    \caption{VMA coefficients for GDP.\\Estimation with pandemic (left) and without pandemic (right).}
\end{figure}

@@

In this section, we demonstrate how pre-filtering indicator variables may improve the performance of simple forecasts by removing noise. We demonstrate how in-sample forecast performance improves when we ``de-noise'' the indicators prior to projection with simple smoothing filters.\footnote{The smoothing filters are one-sided (concurrent) filters of length 31, therefore, estimation of both forecasting models omits the first 30 observations.}

@@

\begin{equation}
    y_{t+h} = \boldsymbol{\beta'} \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $y_{t+h}$ is the quarterly growth rate of GDP for forecast horizon $h = \{0, 1, 2, ... \}$ (labeled shifted GDP in the following )
    \item $\mathbf{X}_t \equiv [1,y_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $y_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\boldsymbol{\beta}}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

@@

\begin{figure}[htbp]

 \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
             \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_all.pdf}
            \caption{Full sample information (without pandemic)\label{direct_wc}}
        \end{center}
    \end{subfigure}

 \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
            \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_financial_crisis.pdf}
            \caption{Financial crisis (without pandemic)\label{direct_wc_financial_crisis}}
        \end{center}
    \end{subfigure}
    \caption{Direct forecasts (red lines) based on all indicators and target (black line): GDP left-shifted  by 0 - 3 quarters.}
\end{figure}

@@

The results indicate that as the forecast horizon $h$ increases, forecasts become less volatile and their turning points (i.e., peaks and troughs) become increasingly right-shifted (lagged) relative to the target GDP. Additionally, the p-values of the HAC-adjusted Wald test precipitously increase from $h=2$ to $h=3$ (see Table \ref{tab:pvaluedhp}), implying that the indicators become statistically insignificant for forecast horizons of more than two quarters.\footnote{ 
These findings are confirmed by the relative root mean-square errors comparing direct forecasts against the mean benchmark, see Table \ref{rRMSE_mSSA_direct_mean_without_covid8} in the Appendix.}

@@

%\begin{table}[ht]
%\centering
%\begin{tabular}{rrrrr}
%  \hline
% & h=3 & h=4 & h=5 & h=6 \\ 
%  \hline
%rRMSE & 0.962 & 0.967 & 0.990 & 0.967 \\ 
%   \hline
%\end{tabular}
%\caption{Relative root mean-square error of HP-C against classic direct forecast for $h\geq %3$.\\Values $<1$ imply  HP-C has smaller RMSE.  } 
%\label{tab:rRMSE}
%\end{table}
\bigskip
\begin{table}[h]
\caption{p-values for $H_0: \boldsymbol{\beta} = 0$ in eq.~\eqref{eq:direct_forecast} based on HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}

@@

\begin{figure}[h]
    \begin{center}
%    \textcolor{blue}{\textbf{Notice that the title on top right panel is incorrect!}}
        \includegraphics[height=4in, width=5.5in]{./Figures/hp_160.pdf}
        % \caption{Two-sided HP(160) (top left) and one-sided (concurrent) HP-C filters (top right) with corresponding amplitude functions (bottom).
        \caption{HP$(\lambda=160)$ filters.\\
        Left: HP (Two-sided, symmetric) \qquad Right: HP-C (One-sided, concurrent)\\
        Top Row: Filter Weights \qquad \qquad \quad Bottom Row: Amplitude Functions.
        \label{hp_160}}
    \end{center}
\end{figure}

@@

Figures \ref{direct_hp_forecasts} and \ref{direct_hp_forecasts_financial_crisis} provide results comparable to those of Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} but now using indicators smoothed with the one-sided HP(160), (denoted as the HP-C, for \textit{concurrent} filter).  The fitted values lag GDP slightly less and track turning points somewhat more quickly. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients ${\boldsymbol{\beta}} = 0$. The use of filtered indicators improves their statistical significance at forecast horizons $h > 2$ quarters. 
%Figures \ref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \ref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic (unfiltered) direct forecasts in the latter figure %in Fig.\eqref{direct_wc_financial_crisis} 
% indicates that the new predictors are slightly less retarded and track the turning points more promptly (`faster'). 
% The filtered approach also exhibits superior statistical properties. As shown in Table \ref{tab:f_stats}, 
%The p-values of the F-statistics for the HP-based forecasts are smaller than for the direct forecasts specifically at $h=3$, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators).

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.85\textwidth]{./Figures/direct_hp_forecasts.pdf}
        \caption{Direct HP-C forecasts: entire data set (without pandemic).
        \label{direct_hp_forecasts}}
    \end{center}
\end{figure}

@@

The application of the one-sided HP-C filter to the indicators yields some measurable improvements in forecast performance. However, 
%its univariate nature excludes additional information such as that provided by leading indicators. 
the noise suppression capability of the one-sided filter is compromised, as explained by \cite{Wildi2025} and illustrated by the high-frequency leakage of the amplitude function in Figure \ref{hp_160}. To address these limitations, we turn to the Multivariate Smooth Sign Accuracy (M-SSA) framework proposed by \cite{Wildi2025}.

@@

Let $\mathbf{x}_t$ be a time series representing the data, with observations $(x_{t},...,x_{1})'$. Let $\mathbf{z}_t=(z_{t},...,z_1)'$  denote a target series, which depends on $\mathbf{x}_t$ as well as on unknown future $x_{t+k}$, $k>0$ (and possibly unknown past $x_k$, $k\leq 0$). For simplicity, we assume stationarity of all time series involved. In our current framework, $z_t$ corresponds to the output of a two-sided Hodrick–Prescott filter with smoothing parameter $\lambda = 160$ applied to the doubly infinite sequence $x_k$, $k\in\mathbb{Z}$. The finite sample $\mathbf{x}_t$ is the portion effectively observed and represents the log-differenced GDP (the dependence on $x_0, x_{-1},...$ is neglected because the HP filter weights decay rapidly to zero; see Figure \ref{hp_160}). The prediction task comprises tracking $z_{t}$  using the (causal) predictor $y_{t}=\sum_{k=0}^{L-1} b_{k}x_{t-k}=\mathbf{b}'\mathbf{x}_t$, where the filter $\mathbf{b}$ with weights $b_0,...,b_{L-1}'$ (actually of length $L=t$) is available for optimization purposes.

@@

Traditional predictors often yield an excessive number of false sign changes due to leakage of high-frequency components (high-frequency leakage) in the predicted series, as discussed above (see Figure \ref{hp_160}). This noisiness is particularly problematic near the onset or termination of economic crises (recessions), where the target fluctuates around zero; reliable tracking of sign changes would facilitate real-time assessment of the occurrence and timing of such events. 
%as illustrated by the amplitude function in Figure \ref{hp_160} (bottom right panel) and further elaborated upon by \cite{Wildi2025}. 
 \cite{Wildi2025} shows that the SSA criterion helps control this phenomenon (Figure \ref{mssa_msse_zc}).

 Using the dual formulation of the criterion \eqref{critssaht}, the author demonstrates that the SSA solution yields the \emph{lowest} zero-crossing rate among all (linear) predictors with equivalent target correlation. 
The concept can be extended to a multivariate framework, denoted M-SSA, by generalizing the correlation functions as detailed in the next section. As we shall see, controlling the zero-crossing rate within the M-SSA framework will have a quantifiable effect on the hit rate and false positive rate, thereby explicitly addressing the detection of present (nowcasting) and future (forecasting) economic contractions and expansions.

We now briefly elucidate the SSA optimization criterion outlined in Eq.\eqref{critssa}. Consider a simplified univariate framework where $x_t=\epsilon_t$ represents white noise with variance $\sigma^2$ and assume, for notational simplicity, that $\sigma^2=1$. A generic target specification for $z_t$ is then defined as a linear filter of $x_t$:
\[z_t=\sum_{k=-\infty}^\infty \gamma_k x_{t-k}=\sum_{k=-\infty}^\infty \gamma_k \epsilon_{t-k},\] 
where it is assumed that $\gamma_k\neq 0$ for some $k<0$ (implying an acausal filter), and $\sum_{|k|<\infty}\gamma_k^2<\infty$ (squared summability), ensuring the stationarity of the target $z_t$. With these definitions, all processes are centered at zero, justifying the use of Equation \eqref{critssaht} which links first-order ACF and HT. Extensions to non-stationary integrated processes are discussed in Wildi (2025). Suppose the goal is to predict  $z_{t+h}$, $h\geq 0$. Let  $y_{h t}=\sum_{k=0}^{L-1} b_{h k}x_{t-k}$  denote a causal predictor of length $L$. For notational convenience, we now omit the subscript $h$ from the predictor $\mathbf{b}=\mathbf{b}_h$ and we denote $\boldsymbol{\gamma}_h:=(\gamma_{h},\gamma_{h+1},...,\gamma_{h+L-1})'$. % To simplify the exposition, we consider the case $h=0$ (nowcast), and thus omit the subscript $h$ from the notation. 
Then, under the postulated white noise assumption, we obtain:
\rho(z,y)=\frac{\mathbf{b}'\boldsymbol{\gamma}_h}{\sqrt{\mathbf{b}'\mathbf{b}}\sqrt{\boldsymbol{\gamma}_{\infty}'\boldsymbol{\gamma}_{\infty}}} \textrm{~and~}\rho(y,1)=\frac{\mathbf{b}'\mathbf{Mb}}{\mathbf{b}'\mathbf{b}},
where $\boldsymbol{\gamma}_{\infty}'\boldsymbol{\gamma}_{\infty}:=\sum_{|k|<\infty}\gamma_k^2<\infty$ and where the so-called autocovariance generating matrix $\mathbf{M}$, of dimension $L\times L$, is defined as:
\mathbf{M}=\left(\begin{array}{ccccccccc}0&0.5&0&0&0&...&0&0&0\\
0.5&0&0.5&0&0&...&0&0&0\\
...&&&&&&&&\\
0&0&0&0&0&...&0.5&0&0.5\\
0&0&0&0&0&...&0&0.5&0
\end{array}\right).
The matrix $\mathbf{M}$ is such that $\mathbf{b}'\mathbf{Mb}=\sum_{k=1}^{L-1}b_{k-1}b_k$, representing the first-order autocovariance of $y_t$.  Using this, the optimization criterion \eqref{critssa} can be reformulated as:
\begin{eqnarray}\label{crit1}
\left.\begin{array}{cc}
&\max_{\mathbf{b}}\mathbf{b}'\boldsymbol{\gamma}_{h}\\
\textrm{s.t.~}&\mathbf{b}'\mathbf{Mb}=l\rho_1\\
&\mathbf{b}'\mathbf{b}=l
\end{array}\right\},
where the additional length constraint $\mathbf{b}'\mathbf{b}=l$ serves two purposes: firstly, to ensure the uniqueness of the SSA solution; secondly, to normalize the objective function $\mathbf{b}'\boldsymbol{\gamma}_{h}$, making it proportional to the target correlation $\rho(z,y)$, up to a fixed scaling factor $\sqrt{\mathbf{b}'\mathbf{b}}\sqrt{\boldsymbol{\gamma}_{\infty}'\boldsymbol{\gamma}_{\infty}}=\sqrt{l}\sqrt{\boldsymbol{\gamma}_{\infty}'\boldsymbol{\gamma}_{\infty}}$. For convenience, $l$ is typically set to 1, yielding a standardized $y_t$. Consequently, the solution to the criterion is invariant to the choice between these two objective formulations. However, 
Criterion \eqref{crit1} is readily amenable to optimization, as it involves affine and quadratic functions of $\mathbf{b}$ \citep{Wildi2024}.\footnote{Detailed derivations are obtained in \cite{Wildi2025}.} \\
Under the above assumptions, the classical MSE predictor of length $L$ can be expressed via orthogonal projection as:
y_{MSE,t}=\sum_{k=0}^{L-1}\gamma_{k+h}x_{t-k}
which corresponds to setting $b_{MSE,k}=\gamma_{k+h}$, $k=0,...,L-1$. In general, to enhance the smoothness of the SSA predictor $y_t$, we select $\rho_1>\rho_{MSE}$ in criterion \eqref{crit1},  where $\rho_{MSE}$ denotes the first-order ACF of the (benchmark) MSE predictor.\footnote{Selecting $\rho_1<\rho_{MSE}$ would produce an SSA predictor with more frequent zero-crossings than the MSE benchmark, which is typically undesirable in practical applications.} Finally,  an extension to autocorrelated $x_t\neq \epsilon_t$ is discussed below.
